### 2013-10-14 writeups/context-dependence.tex
.

  \begin{center}
    \begin{tabular}{c@{\quad$\to$\quad}c@{\quad at rate\quad }c}
      $u$  &  $v$  &  $\mu$  \\
      \hline
      $+++$  &   $+-+$   &  $\lambda/1+e^{2\beta - \gamma})$ \\
      $++-$  &   $+--$   &  $\lambda/(1+e^{-\gamma})$ \\
      $+-+$  &   $+++$   &  $\lambda/(1+e^{-2\beta + \gamma})$ \\
      $+--$  &   $++-$   &  $\lambda/(1+e^{\gamma})$ \\
      $-++$  &   $--+$   &  $\lambda/(1+e^{-\gamma})$ \\
      $-+-$  &   $---$   &  $\lambda/(1+e^{-2\beta - \gamma})$ \\
      $--+$  &   $-++$   &  $\lambda/(1+e^{\gamma})$ \\
      $---$  &   $-+-$   &  $\lambda/(1+e^{2\beta + \gamma})$ 
    \end{tabular}
  \end{center}

@@

We clearly don't want to add an extra parameter $\pi(z)$ for all $z \in \calS^{m+2\ell}$. 
If nothing else, it would make the inference depend on $m$ and $\ell$, which we'd like to avoid.
One approach would be to put a Dirichlet distribution on $\pi(z)$.
This has the nice property that you can choose priors that are consistent across different values of $m$ and $\ell$.
We'd have to still MCMC around in the space of $\pi(z)$, thought, so they'd be nuisance parameters from that point of view.
Also, making the frequencies at the root random introduces correlations between counts, which we want to avoid.
An alternative is to specify $\pi$ through lower-dimensional statistics,
for instance to model the root state as a sample from a Markov random field
with parameters for the marginal frequencies of each state, and pairwise (or higher) correlations between them.

A Markov random field with interactions between sites no more than distance two away 
is determined by the energy function $\phi(a,b,c)$ which gives the ``energy'' of the configuration $abc$ for each $a,b,c \in \calS$.
We use this to define the distribution on root patterns $\pi$ by
\begin{align}  \label{eqn:root_mrf}
    \pi(z) = \frac{1}{Z} \exp\left( - \sum_{i=1}^{|z|} h(z_{i-1},z_i,z_{i+1}) \right),
\end{align}
i.e.\ the exponential of the total energy of each three adjacent sites,
where $Z$ is the normalizing constant.
Note that it is often difficult to compute $Z$ in applications of Markov random fields,
but here we are working with relatively short sequences, 
so we just compute \eqref{eqn:root_mrf} for all possible $z$, and divide by the sum.
(However, for this same reason this is not equivalent to saying that $X(0)$ is a sample from the Markov random field
and finding the distribution of patterns within this.)

@@

  \item Compute $M_4(x,(y,z)) = \sum_{v \in \calS^{m+2\ell}} p_{m+2\ell}(t_{A},v,x) M_3(v,z) M_4(v,y)$ (stored as a $|\calS|^{m+2\ell} \times |\calS|^{2m}$ matrix).

@@

Strings of length $10^6$ were evolved under Glauber dynamics on the Ising model with varying parameters,
and pattern occurrences counted.
As for TASEP, unning 10,000 MCMC iterations with window sizes of $\ell=w=3$ (total length 9) on these takes around 40 minutes.
In this case, the true parameter values have good distribution within the estimated posterior distributions.

@@

>>>>

  \begin{center}
    \begin{tabular}{c@{\quad$\to$\quad}c@{\quad at rate\quad }c}
      $u$  &  $v$  &  $\mu$  \\
      \hline
      $+++$  &   $+-+$   &  $\lambda/(1+e^{2\beta - \gamma})$ \\
      $++-$  &   $+--$   &  $\lambda/(1+e^{-\gamma})$ \\
      $+-+$  &   $+++$   &  $\lambda/(1+e^{-2\beta + \gamma})$ \\
      $+--$  &   $++-$   &  $\lambda/(1+e^{\gamma})$ \\
      $-++$  &   $--+$   &  $\lambda/(1+e^{-\gamma})$ \\
      $-+-$  &   $---$   &  $\lambda/(1+e^{-2\beta - \gamma})$ \\
      $--+$  &   $-++$   &  $\lambda/(1+e^{\gamma})$ \\
      $---$  &   $-+-$   &  $\lambda/(1+e^{2\beta + \gamma})$ 
    \end{tabular}
  \end{center}

@@

We clearly don't want to add an extra parameter $\pi(z)$ for all $z \in \calS^{m+2\ell}$. 
If nothing else, it would make the inference depend on $m$ and $\ell$, which we'd like to avoid.
One approach would be to put a Dirichlet distribution on $\pi(z)$.
This has the nice property that you can choose priors that are consistent across different values of $m$ and $\ell$.
We'd have to still MCMC around in the space of $\pi(z)$, so they'd be nuisance parameters from that point of view.
Also, making the frequencies at the root random introduces correlations between counts, which we want to avoid.
An alternative is to specify $\pi$ through lower-dimensional statistics,
for instance to model the root state as a sample from a Markov random field
with parameters for the marginal frequencies of each state, and pairwise (or higher) correlations between them.

A Markov random field with interactions between sites no more than distance two away 
is determined by the energy function $\phi(a,b,c)$ which gives the ``energy'' of the configuration $abc$ for each $a,b,c \in \calS$.
We use this to define the distribution on root patterns $\pi$ by
\begin{align}  \label{eqn:root_mrf}
    \pi(z) = \frac{1}{Z} \exp\left( - \sum_{i=1}^{|z|} \phi(z_{i-1},z_i,z_{i+1}) \right),
\end{align}
i.e.\ the exponential of the total energy of each three adjacent sites,
where $Z$ is the normalizing constant.
Note that it is often difficult to compute $Z$ in applications of Markov random fields,
but here we are working with relatively short sequences, 
so we just compute \eqref{eqn:root_mrf} for all possible $z$, and divide by the sum.
(However, for this same reason this is not equivalent to saying that $X(0)$ is a sample from the Markov random field
and finding the distribution of patterns within this.)

@@

  \item Compute $M_4(x,(y,z)) = \sum_{v \in \calS^{m+2\ell}} p_{m+2\ell}(t_{A},v,x) M_2(v,z) M_3(v,y)$ (stored as a $|\calS|^{m+2\ell} \times |\calS|^{2m}$ matrix).

@@

Strings of length $10^6$ were evolved under Glauber dynamics on the Ising model with varying parameters,
and pattern occurrences counted.
As for TASEP, running 10,000 MCMC iterations with window sizes of $\ell=w=3$ (total length 9) on these takes around 40 minutes.
In this case, the true parameter values have good distribution within the estimated posterior distributions.

@@

sergey:
\begin{itemize}

  \item ancestral CpG

  \item ancient DNA
